Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap
Half a page

Because we have a denotational semantics, it is easy to validate equations
between terms, blocks, and CFGs.

Even the basic proof of the soundness of substitution already lets us prove the
soundness of many fundamental compiler optimisations, such as global value
numbering (let-introduction for pure expressions), some forms of dead code
elimination (i.e., eliminating unused affine bindings), and strength reduction
with loop hoisting (let-introduction for relevant expressions).

@@

We will begin by giving a useful foundation for a wide class of \isotopessa
models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially
infinite) nondeterministic trace of monoidal "events". These will later come in
handy for reasoning about output behaviour and weak memory. We start with some
definitions:
\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a
  monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to
  I\) mapping (infinite) streams of \(M\) to elements of \(I\) satisfying
  \(
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the
  stream \(\ell\).
  Some basic examples include:
  \begin{enumerate}
    \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
    forms a stream action on the unit set.
    % \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
    % \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
    % \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
    % \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
    % \cdot_i l_i)_i\).
    \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a
    stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq
    \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\),
    where \(\ell \cdot s\) prepends the \(\ell\) to \(s\).
    % \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
    % stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
    % on \(\nats \cup \{\infty\}\) by addition.
    \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
    stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
    \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
    compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
  \end{enumerate}
\end{definition}
% We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
% \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
% those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
% For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
% \omega}) = \{[\,]\}\). Generally, \(\mc{SZ}(M, I)\) always contains
% the unit of \(M\). 
Using stream actions to determine what it means to perform a monoidal action
"infinitely many times," we can now define the \textit{trace monad} of a stream
action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta: M^\omega \to I\), we can define the
  \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
  \times M + I)\) with unit \(\eta_A = \lambda a.\iota_0 (a, 1)\) and
  multiplication
  \(
    \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
  \),
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). We can
  equip this monad with an iteration operator \(f^\dagger(a) = f^\infty(a) \cup
  \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
  \in f_n(a)\}\), yielding an Elgot structure, where
  \[
    f_0 = f, \quad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n], \quad
    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
  \]
\end{definition}

We now give a few basic definitions relevant to monads over \(\mb{Set}\):
\begin{definition}[Relevant, Affine Monad]
  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
  if it is both relevant and affine.
\end{definition}
\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
  from \(\mb{T}\). Some important examples of submonads include:
  \begin{enumerate}
    \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
    \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
    %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
    \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
    intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
    submonads, forming a complete meet-semilattice with bottom element \(\eta\)
    and top element \(\mb{T}\),
    % ; this induces a complete lattice in the standard manner.
    \item In particular, the intersection of a relevant/affine/commutative
    submonad and another submonad is relevant/affine/commutative respectively.
  \end{enumerate}
\end{definition} 
We can use the trace monad described above to derive some interesting examples
of submonads:
\begin{enumerate}
  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
  \ms{Traces}\;\Theta\;A\) are submonads
  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
  a submonad whose Kleisli category is a central subcategory of the Kleisli
  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
  submonoid.
  \label{item:central-submonad}
  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
  \label{item:relevant-submonad}
  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
  \label{item:affine-submonad}
\end{enumerate}
Since every commutative submonad induces a central subcategory of the Kleisli
category of the parent monad, it follows that a triple of a relevant, affine,
and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into a
substructural effectful category with all objects both affine and relevant. In
particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
\ref{item:affine-submonad} equip the Kleisli category of
\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
category given choices of \(M_C\), \(M_R\).

@@

As a concrete example, we can now introduce the \textbf{printing monad}, which
is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
\(\Sigma: (\ms{Byte}^*)^\omega \to \ms{Byte}^{\leq \omega}\) denotes
concatenation of bytestrings into a potentially infinite bytestream. This can be
equipped with the structure of a substructural effectful category by choosing
\(M_C = M_R = \{[]\}\), giving us an \isotopessa model supporting the
instructions \(\ms{print} \in \mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and
\(\ms{nondet} \in \mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard
semantics \(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\),
\(\dnt{\ms{nondet}} = \lambda (). \{\iota_0 (a, []) \mid a \in A\}\).

We can make things a little more interesting by adding a heap to the mix,
defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The
Kleisli category of this monad can be equipped with the structure of a
substructural effectful category by simply taking the affine/relevant morphisms
to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\);
as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we
therefore have an \isotopessa model supporting the instructions \(\ms{set}:
\mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
\mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats,
\nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign
semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} =
\lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit
more tricky, since it is unclear what to do when we try to access uninitialized
memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\); for now,
this will do. Finally, \(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', [])
\mid h' = h \sqcup p \mapsto v\}\) simply fills a random empty heap cell with
the provided value, returning a pointer to the cell, while \(\dnt{\ms{free}} =
\lambda p\;h. \{\iota_0\;((), h \setminus p, [])\}\)

We can use this construction to, building on the ideas in \cite{mellies-ftrs},
show how a simple separation logic based on \cite{reynolds-separation-2002}
slightly adapted to the categorical setting gives us a very rich \isotopessa
model supporting a basic notion of refinement types, while at the same time
giving us a nontrivial notion of nonlinear types.

Given a set \(A\), we can define a heap-predicate \(\Phi\) over \(A\) to be of
type \(A \to \ms{Heap} \to \ms{Prop}\). Taking as objects pairs \((A, \Phi)\),
we can define the category \(\ms{Sep}\) to have morphisms

@@

is the usual partial correctness triple definition and the separating
conjunction of heap-predicates is defined as usual to be 

@@

\Psi) = (A \times B, \Phi \ast \Psi)\), verifying that, taking

@@

equips \(\ms{Sep}\) with the structure of a premonoidal category. We can give \(\ms{Sep}\) the structure of a substructural effectful category by taking:
  \item Affine objects to be precisely those for which \(\ms{drop} =
  \upg{\lambda a. ()}{} \in \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), and
  relevant objects to be precisely those for which \(\ms{split} = \upg{\lambda
  a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi * \Phi))\), which for this logic
  corresponds to pairs \((A, \Phi)\) where \(\Phi\) does not depend on the heap.

@@

Another interesting use of the trace monad to build a sophisticated \isotopessa
model is to build a model of TSO-style weak memory based on \citet{sparky}. For
simplicity, we will consider an infinite set of named locations \(x, y, z \in
\ms{Loc}\) which are subject to concurrent modification by all threads via TSO
reads, writes, and fences. We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
  from \(P\) and \(Q\), there exists a partial order isomorphism between them
  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
  if its carrier set is. We will write unordered pomsets using multiset
  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
  notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
predecessors. They also form a monoid under parallel composition \(\alpha ||
\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
since we only consider nonempty pomsets but allow the removal of finitely many
ticks.

Given a partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n
\in N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by
the function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega
\to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on
finite pomsets.

We define a \textbf{program order pomset} to be a pomset with \(\mc{A}_{\ms{PO}}
= \mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is taken as a stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with semantics
\begin{equation}
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

>>>>

Because we have a denotational semantics, it is easy to validate equations
between terms, blocks, and CFGs. Even the basic proof of the soundness of
substitution already lets us prove the soundness of many fundamental compiler
optimisations, such as global value numbering (let-introduction for pure
expressions), some forms of dead code elimination (i.e., eliminating unused
affine bindings), and strength reduction with loop hoisting (let-introduction
for relevant expressions).

@@

We will begin by giving a useful foundation for a wide class of \isotopessa
models: the "trace models" over \(\mb{Set}\), which accumulate a (potentially
infinite) nondeterministic trace of monoidal "events". These will later come in
handy for reasoning about output behaviour and weak memory. We start with some
definitions:
\begin{definition}[Relevant, Affine Monad]
  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
  if it is both relevant and affine.
\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
  from \(\mb{T}\). Some important examples of submonads include:
  \begin{enumerate}
    \item The \textbf{unit submonad} \(\eta\) of every monad \(\mb{T}\) with
    \(\eta(A) = \{\eta_A\;a \mid a \in A\}\).
    %; similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\). 
    \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
    intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
    submonads, forming a complete meet-semilattice with bottom \(\eta\) and top
    \(\mb{T}\),
    % ; this induces a complete lattice in the standard manner.
    \item In particular, the intersection of a relevant/affine/commutative
    submonad and another submonad is relevant/affine/commutative respectively.
  \end{enumerate}
\begin{definition}[Stream Action]
  A \textbf{stream action} of a monoid \(M\) on a set \(I\) equipped with a
  monoid action \(\cdot: M \times I \to I\) is a function \(\Theta: M^\omega \to
  I\) mapping (infinite) streams of \(M\) to elements of \(I\) satisfying
    m \cdot \Theta(\ell) = \Theta(m \colon \ell)
  \),
  where \(m \colon \ell\) denotes the stream formed by prefixing \(m\) to the
  stream \(\ell\).
  Some basic examples include:
  \begin{enumerate}
    \item For any monoid \(M\), the unique map \(\Theta: M^\omega \to \mathbf{1}\)
    forms a stream action on the unit set.
    % \item In particular, given monoids \(M_i\) acting on sets \(I_i\) with action
    % \(\cdot_i\) and stream actions \(\Theta_i: M_i^\omega \to I_i\), \(\Theta =
    % \langle \Theta_i \circ \ms{fmap}\;\pi_i\rangle_i: (\Pi_iM_i)^\omega \to
    % \Pi_iI_i\) is a stream action compatible with the action \(m \cdot l = (m_i
    % \cdot_i l_i)_i\).
    \item The concatenation map \(\Theta: (A^*)^\omega \to A^\omega\) is a
    stream action of \(A^*\), the set of (finite) lists of \(A\), on \(A^{\leq
    \omega}\), the set of (\textit{potentially} infinite) sequences of \(A\),
    where \(\ell \cdot s\) prepends the \(\ell\) to \(s\).
    % \item The sum map \(\Sigma: \nats^\omega \to \nats \cup \{\infty\}\) is a
    % stream action of \(\nats\) on \(\nats \cup \{\infty\}\), where \(\nats\) acts
    % on \(\nats \cup \{\infty\}\) by addition.
    \item Given a monoid \(M\) acting on sets \(I_i\) with action \(\cdot_i\) and
    stream actions \(\Theta_i: M^\omega \to I_i\), \(\Theta =
    \langle\Theta_i\rangle_i: M^\omega \to \Pi_iI_i\) is a stream action
    compatible with the action \(m \cdot l = (m \cdot_i l_i)_i\).
  \end{enumerate}
% We define the \textbf{stream center} of \(M\) with respect to \(I\), denoted
% \(\mc{SZ}(M, I)\), as the subset of the center of \(M\), \(Z(M)\), consisting of
% those elements \(m \in Z(M)\) such that for all \(i \in I\), \(m \cdot i = i\).
% For instance, \(\mc{SZ}(M, \mathbf{1}) = Z(M)\), and \(\mc{SZ}(A^*, A^{\leq
% \omega}) = \{[\,]\}\). Generally, \(\mc{SZ}(M, I)\) always contains
% the unit of \(M\). 
Using stream actions to determine what it means to perform a monoidal action
"infinitely many times," we can now define the \textit{trace monad} of a stream
action as follows:
\begin{definition}[Trace Monad]
  Given a stream action \(\Theta: M^\omega \to I\), we can define the
  \textbf{nondeterministic trace monad} \(\ms{Traces}\;\Theta\;A = \mc{P}^+(A
  \times M + I)\) with unit \(\eta_A = \lambda a.\iota_0 (a, 1)\) and
  multiplication
    \mu_A\;T = \{[\lambda (a, m). m \cdot a, \ms{id}](t) \mid t \in T\}
  \),
  where the action of \(M\) on traces is given by \(m \cdot \iota_0 (a, m') =
  \iota_0 (a, mm')\) and \(m \cdot \iota_1 t = \iota_1 (m \cdot t)\). We can
  equip this monad with an iteration operator \(f^\dagger(a) = f^\infty(a) \cup
  \bigcup_{n \in \nats}\{\iota_0 (\iota_0\;b, m) \in f_n(a)\} \cup \{\iota_1 t
  \in f_n(a)\}\), yielding an Elgot structure, where
  \[
    f_0 = f, \quad f_{n + 1} = f \gg [\upg{\iota_0}{}, f_n], \quad
    f^\infty(a) = \{\Theta (\lambda n. m_n) \mid a_0 = a \land \forall n, \exists a_{n + 1}. \iota_0 (\iota_1\;a_{n + 1}, m_n) \in f(a_n)\}
  \]
We can use the trace monad described above to derive some interesting examples
of submonads:
\begin{enumerate}
  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
  \ms{Traces}\;\Theta\;A\) are submonads
  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
  a submonad whose Kleisli category is a central subcategory of the Kleisli
  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
  submonoid.
  \label{item:central-submonad}
  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
  \label{item:relevant-submonad}
  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
  \label{item:affine-submonad}
\end{enumerate}
Since every commutative submonad induces a central subcategory of the Kleisli
category of the parent monad, it follows that a triple of a relevant, affine,
and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into a
substructural effectful category with all objects both affine and relevant. In
particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
\ref{item:affine-submonad} equip the Kleisli category of
\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
category given choices of \(M_C\), \(M_R\).

@@

We can now introduce the \textbf{printing monad}, which is simply defined as
\(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where \(\Sigma:
(\ms{Byte}^*)^\omega \to \ms{Byte}^{\leq \omega}\) denotes concatenation of
bytestrings into a potentially infinite bytestream. This can be made into a
substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving us
an \isotopessa model supporting instructions \(\ms{print} \in
\mc{I}^\varnothing_0(\ms{Byte}^*, \mb{1})\) and \(\ms{nondet} \in
\mc{I}^\taff_1(\mb{1}, A)\) (for \(A\) nonempty), with semantics
\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), b)\}\), \(\dnt{\ms{nondet}} =
\lambda (). \{\iota_0 (a, []) \mid a \in A\}\).

We can make things a little more interesting by adding a heap to the mix,
defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where
\(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function with
finite support. The Kleisli category of this monad is a substructural effectful
category and inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\);
we therefore have an \isotopessa model supporting the instructions \(\ms{set}:
\mc{I}^\varnothing_0(\nats \times \nats, \mb{1})\), \(\ms{get}:
\mc{I}^\varnothing_0(\nats, \nats)\), \(\ms{alloc}: \mc{I}^\varnothing_0(\nats,
\nats)\), and \(\ms{free}: \mc{I}^\varnothing_0(\nats, \mb{1})\). We can assign
semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} =
\lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit
more tricky, since it is unclear what to do when we try to access uninitialized
memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}}
= \lambda p\;h. \{\iota_0(v, h, []) \mid v = h\;p \lor p \notin h\}\). Finally,
\(\dnt{\ms{alloc}} = \lambda v\;h.\{\iota_0\;(p, h', []) \mid h' = h \sqcup p
\mapsto v\}\) simply fills a random empty heap cell with the provided value,
returning a pointer to the cell, while \(\dnt{\ms{free}} = \lambda p\;h.
\{\iota_0\;((), h \setminus p, [])\}\).

Building on the ideas in \citet{mellies-ftrs}, we show how a simple separation
logic based on \citet{reynolds-separation-2002} slightly adapted to the
categorical setting gives us a very rich \isotopessa model supporting a basic
notion of refinement types, while at the same time giving us a nontrivial notion
of nonlinear types. Taking as objects pairs \((A, \Phi)\), where \(\Phi: A \to
\ms{Heap} \to \ms{Prop}\), we can define the category \(\ms{Sep}\) to have
morphisms

@@

corresponds to the usual notion of a partial correctness triple, and the
separating conjunction of heap-predicates is defined as

@@

\Psi) = (A \times B, \Phi \ast \Psi)\), making \(\ms{Sep}\) into a substructural
effectful category with:

@@

  \item Affine objects those for which \(\ms{drop} = \upg{\lambda a. ()}{} \in
  \ms{Sep}((A, \Phi), (\mb{1}, \ms{emp}))\), and relevant objects those for
  which \(\ms{split} = \upg{\lambda a. (a, a)} \in \ms{Sep}((A, \Phi), (A, \Phi
  * \Phi))\), which for this logic means pairs \((A, \Phi)\) where \(\Phi\) does
  not depend on the heap.

@@

Another interesting use of the trace monad to build a sophisticated \isotopessa
model is to build a model of TSO-style weak memory based on \citet{sparky}. For
simplicity, we will consider an infinite set of named locations \(x, y, z \in
\ms{Loc}\) which are subject to concurrent modification by all threads via TSO
reads, writes, and fences. We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
  \to \mc{A}\) quotiented over the removal of finitely many \(\delta\). A pomset
  is \textbf{finite} if its carrier set is. We will write unordered pomsets
  using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets
  using list notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
predecessors. They also form a monoid under parallel composition, defining
\(\alpha || \beta = [\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given
the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).
Note in both cases the monoidal unit is \(\{\delta\}\), since we only consider
nonempty pomsets but allow the removal of finitely many ticks. Given any
partially ordered set \(N\) and a family of pomsets \(\alpha_n\) for \(n \in
N\), we can define their \textit{sum} \(\Sigma_n\alpha_n\) to be given by the
function \((n, a) \mapsto \alpha_n\;a: \ms{trim}(\Sigma_nP_n) \to \mc{A}\),
where the dependent product \(\Sigma_nP_n\) is given the lexicographic order. In
particular, choosing \(N = \nats\) makes \(\Sigma: \ms{Pom}_{\ms{fin}}^\omega
\to \ms{Pom}\) into a stream action w.r.t the sequential composition monoid on
finite pomsets.
